---
title: "LDAPlots"
author: "James Colee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LDAPlots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Introduction

The LDAPlots package was created to help with the interpretations of multivariate analysis resulting from the problem of having multiple response variables and a single categorical predictor variable. Some of the common ways of dealing with these situations are through multivariate analysis of variance (MANOVA), analysis of similarities (ANOSIM), or permutational multivariate analysis of variance (PERMANOVA). The problem with all these methods is that there is not an easy way to visualize these results when the categorical variable has more than three levels. Descriptions of these types of analyses, functions, and how to perform them can be found in the package  [vegan](https://cran.r-project.org/web/packages/vegan/vegan.pdf).

```{r setup}
library(LDAPlots)
```

## LDA Projection Plots

To allow for better interpretation and visualization of these types of multivariate results, a linear discriminant analysis (LDA) can be performed. LDA can project k-dimensional space into 2-dimensional space. The projected space can be plotted as a 2-dimensional plot. The function `lda_plot()` plots the resulting rotation of points along with the Normal 95% data ellipses. The input is an lda class object resulting from `lda()` in the package `MASS` and a vector of the group identifications for each row of data in the original data set. It returns a gglot object, which can be modified just as other ggplot objects can. Examples of this are shown in the confidence intervals section. Along with the rotations, the graphs can be modified to include group means, biplots, and three difference types of mean confidence intervals.



```{r fig.height = 5, fig.width = 6, fig.align = "center"}
library(MASS)
iris.lda <- lda(Species ~., data = iris)
p<-lda_gplot(lda=iris.lda, groups = iris$Species)
p
```

**Confidence Intervals**

Three types of confidence bounds can be produced by `lda_gplot`: asymptotic, bootstrap, and permutation. In all three cases, the rotations are assumed to be constant. This means that only the rotated data needs to be bootstrapped or permuted, depending on the desired confidence interval. Keeping the rotations constant makes sense – otherwise, each new LDA rotation would be projected into different 2-dimensional space. 

**Asymptotic Confidence Circles**

The asymptotic confidence circles for the means can be added to the plot with the `circles = TRUE` setting. The confidence circles have a radius of the square root of a chi square, with G degrees of freedom divided by the number of observations in each group.

$$
asymptotic\,\,confidence\,\,circles\,\,radii = \sqrt\frac{\chi_{ G, 1-\alpha } }{n_{g}}
$$


```{r fig.height = 5, fig.width = 6, fig.align = "center"}
library(ggplot2)
p<-lda_gplot(lda=iris.lda, groups = iris$Species,
             normal.CI = FALSE, circles = TRUE, biplot = FALSE)
p + ggtitle("95% asymtopic CI's")
```

**Bootstrap Confidence Intervals**

The bootstrap confidence intervals for the means can be added to the plot with the `boot.CI = TRUE` setting. The bootstrap estimates are obtained via bootstrap resampling of the LDA rotations, calculation of the group means, and estimation of the normal confidence intervals based on simulated means. The bootstrap confidence intervals are a non-parametric approximation of the individual group mean variability.

```{r fig.height = 5, fig.width = 6, fig.align = "center"}
p<-lda_gplot(lda=iris.lda, groups = iris$Species,
             normal.CI = FALSE,boot.CI = TRUE, fill = FALSE, biplot = FALSE)
p + ggtitle("95% bootstrap CI's")
```

**Permutation Confidence Intervals**

The permutation confidence intervals for the means can be added to the plot with 'circles = TRUE'. The permutation error estimates are obtained via permutation of group associations of the LDA rotations. Then, the errors are added back to the original means to give the permuted mean estimates. Finally, the normal confidence intervals are estimated based on the permuted means. The permutation confidence intervals are a non-parametric approximation of the pooled group mean variability.

```{r fig.height = 5, fig.width = 6, fig.align = "center"}
p<-lda_gplot(lda=iris.lda, groups = iris$Species,
             normal.CI = FALSE,perm.CI = TRUE, fill = FALSE, biplot = FALSE)
p + ggtitle("95% permutation CI's")
```


## Posterior Probability Plots

An alternative way to visualize the rotations is to plot the posterior probabilities of group membership on the rotated axis. Posterior probabilities are the predicted probabilities of group membership based on an LDA model. The function `lda_posterior_plot()` plots the resulting rotation of points along with the posterior probabilities. The input is an lda class object resulting from `lda()` in the package `MASS` and a vector of the group identification for each row of data in the original data set. It returns a set of graphs for each group.

```{r fig.height = 5, fig.width = 6, fig.align = "center"}
lda_posterior_plot(lda=iris.lda, groups = iris$Species)
```

## Other functions

Two other functions are included in the package `boot_sample()` and `perm_sample()`. Both functions were originally designed as internal functions but were given user access. They can be used to create bivariate bootstrap and permutation mean samples for data with group associations.

Steps for using the simulation functions:

1. Simulate a data set with random group labels.

```{r}
q=100
data <- as.data.frame(list(x1 = runif(q), x2 = rnorm(q), x3 = rlnorm(q), 
                           group = sample(c('s','d','w'),q,replace=TRUE)))
```

2. Formatting of data with lda(), this step is not require if data is already bivariate.

```{r}
lda <- lda(group ~., data = data)
# format data, groups need to be numeric, can not be factors of characters
V1<-as.numeric(as.factor(data$group))
lda.vec<-as.data.frame(lda$scaling)
lda.p <- predict(lda)
v <- as.data.frame(cbind(V1, lda.p$x))
```

Input data structure.

```{r}
str(v)
```

3. Create bootstrap sample means.

```{r}
b <- boot_sample(values = v)
str(b)
```


3. Create permutation sample means.

```{r}
b <- boot_sample(values = v)
str(b)
```
